Giant Low Surface Brightness Galaxies As Merger Remnants

### 2016-02-03 glsb.tex
ack
fixing plots and text

The another formation scenario for GLSB systems is that they are normal galaxies that undergo a high angular momentum minor merger where the LSB disk is the shredded remains of an accreted galaxy.  \citet{Pen06} find that extended disks can be formed via the disruption of dwarf galaxies.  These simulated debrie disks have exponential profiles and extend out to 200 kpc from the central galaxy.

Observations of Malin~1 support the accretion model of GLSBs.  \citet{Barth07} analyze HST imaging of Malin~1 and find that the central object is well described as a normal early-type barred galaxy that was previously unresolved in ground-based imaging. \citet{Barth07} points out that exteded disk structures in galaxies such as M31 appear to be a smooth continuation of the inner disk \citep{Ferguson02}, while Malin~1's extended disk is better described as a distinct component, (although \citet{Ibata05} claim M31's extended disk is of accretion origin).

If GLSB disks are indeed the remains of accreted dwarf galaxies, they represent a unique opportunity to observationally probe the outer regions of galaxy dark matter halos, especailly if they form around elliptical galaxies.  The mass profiles of elliptical galaxies have been difficult to constrain since they lack the extended \HI\ disks found around spirals.  As an example of how difficult it can be to measure the outer regions of ellipical galaxy kinematics, \citet{Murphy2011} model M87 out to $\sim$23 kpc with stellar kinematics and out to 51 kpc with globular clusters (or $R=4.8 R_e$, where $R_e$ is the effective radius). \citet{Raskutti14} are also limited to measuring kinematics out to $\sim$2-5 $R_e$.

@@

\begin{deluxetable*}{l l c c c c c c}
%\tabletypesize{\small \footnotesize \scriptsize}
\tabletypesize{\small}
%\rotate
\tablewidth{0pt}
%\tablenum{num}
%\tablecolumns{num}
%\tableheadfrac{num}
\tablecaption{Basic Properties of GLSB Galaxies}
\tablehead{
\colhead{Name} & \colhead{SDSS ID}   &  \colhead{z}  &\colhead{$r$}  & \colhead{$R_e$} 
& \colhead{$\sigma$} & Stellar Mass & \colhead{\HI\ mass}\\
     &  \colhead{(DR 9)}   &     &  \colhead{(mag)}    & \colhead{arcsec} & \colhead{\kms}  &\colhead{log$_{10}$M$_{\odot}$}&\colhead{log$_{10}$M$_{\odot}$}
\label{basic_prop}}    
\startdata
XXX-add Malin 1
Malin 2 & 1237667735572381852 &   0.046 & 14.34 & 7.17 & 202  & 11.15 & 10.6\\
UGC 06614 & 1237668585435365393 &  0.021 & 13.52 & 7.3 & 150 & 10.66 & 10.5
\enddata
\tablecomments{\HI\ masses updated to our adopted cosmological parameters from \citet{Pickering97}.  The rest taken from SDSS.}
\end{deluxetable*}

@@

\begin{figure}
\epsscale{.4}
\plotone{Plots/Color_Magnitu_Diagram.eps}\plotone{Plots/Faber_Jackson_Relation.eps}\\
\plotone{Plots/Fundamental_Plane.eps}\plotone{Plots/mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin 2 and UGC 6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band fundamental plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   Looking at the SDSS data, our GLSB galaxies appear as realtively typical red elliptical galaxies.  \label{elliptical_relations}}
\end{figure}

@@

In Figure~\ref{disk_comp}, we compare the sizes of the stellar disks of the GLSB galaxies with other observations.  For this plot, we take the maximum detected radius from the literature for the galaxies.  It is possible that deeper observations could increase the values of $R_{max}$.  While UGC 6614 is close in size to the upper end of local \HI\ disks, Malin 1 and 2 are much larger than galaxies observed in the THINGS survey.  This seems to point against the GLSB formation mechanism of  \citet{Mapelli08}, as we do not observe \HI\ disks large enough to convert into GLSB systems unless an interaciton is able to boost the \HI\ angular momentum.

%/astro/store/shared-scratch1/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, Malin 1 from \citet{Moore06}, Malin 2 and UGC 6614 from \citet{Pickering97}.  Here, $R_{max}$ is the maximum radius with an \HI\ detection for the THINGS galaxies, Malin~2, and UGC~6614 and the maximum radius with a stellar flux detection for Malin~1 XXX.  It is interesting to note that the \emph{stellar} disks of GLSB galaxies can be larger than the \HI\ disks of nearby systems.  \label{disk_comp}}
\end{figure}

@@

We use the kinemetry package presented in \citet{Krajn06} to fit a tilted-ring model to the stellar kinematics.  For these fits, we hold the position angle and inclination fixed to values found from fitting the fiber-photometry.  Figures~\ref{malin_kinem} and~\ref{ugc_kinem} show the kinemetry fits and extracted rotation curves.  We iterate the fit, clipping residuals which deviate from the model my more than 250 \kms.

@@

Following the work in \citet{Yoachim10,Yoachim12}, we generate a suite of model galaxy SEDs using relatively simple star formation histories and find the best fitting single model.  We break the star formation history into three age bins, 0-10 Gyr, 10-11.5 Gyr, and 11.5-12 Gyr.  We assume each age bin has a constant star formation with a level of 0, 0.5, 1, or 10 solar-masses per year.  Each SFH can have a metallicity of [Z/H] = -2.3,-1.70,-0.70, -0.40, 0.0, or  0.40.  We do not include any metallicity evolution, as all the age bins in a given model has the same metallicity.  This gives us 361 unique non-zero model spectra. XXX-cite which SED code I'm using and IMF etc.

@@

%--now on new mac.  /Users/yoachim/Research/GLSB/THINGS/guess_projen.py
\begin{figure}
\epsscale{0.5}
\plotone{Plots/things_projenguess.eps}
\epsscale{1}
\caption{Comparison of GLSB \HI\ masses \citep{Pickering97} with \HI\ masses measured in the THINGS survey \citep{Walter2008}.  While the \HI\ masses in the GLSB systems are high, they are consistent with what would be available in a merger with a V$_{\rm{max}}\sim$110 km/s or smaller galaxy, assuming the merger did not trigger large amounts of star formation and feedback. \label{thingsMHI}}
\end{figure}

@@

@@

We have observed two giant low surface brightness galaxies with the VIRUS-P IFU.  We find the systems are best explained as massive elliptical galaxies which have managed to accrete stars and gas into a large disk.  We find the inner bulge is similar to elliptical galaxies found in SDSS.  The stars have a disk-like velocity field, are dominated by old stars, but do show signs of ongoing star formation. The stars also show a sharp metalicity transition, consistent with the disk forming independently of the bulge. The \HI content of the disk is also consistent with coming from a low-mass accreted galaxy.

@@

\appendix
\section{SDSS SQL Querries} \label{SQL}
To build our sample of comparison galaxies from SDSS, we used the CasJobs interface to query the SDSS DR10 database.

@@

% LocalWords:  Debattista Horrocks Lancashire RCUK Malin Stanonik Spavone Lelli
% LocalWords:  Bothun GLSB Das Mapelli Weij Loebman SungWon Kwak Scd Sd
% LocalWords:  Schiminovich Iodice Belokurov Fairchild RMS pPXF Cappellari th
% LocalWords:  GANDALF Sarzi CosmoloPy Bernardi GLSBs DeBlok Verheijen Yoachim
% LocalWords:  SEDs Brookhaven Instituto Astrofisica Canarias Notre JINA DM
% LocalWords:  IFU UGC HSB SAURON ellipticals Spitzer IC
% LocalWords:  kpc HIPASS CCD metallicity MANGA SDSS arcsec lensing IRAF
% LocalWords:  Spectrograph spectrograph IDL longslit photometric FWHM SSP mis
% LocalWords:  dispersions Balmer decrement fiducia redshift
% LocalWords:  Faber redshifts NGC kinemetry photometry anisotropy Gyr SFH
% LocalWords:  metallicities colliders rotators et al MNRAS de datacube Ppxf
% LocalWords:  dashed CDM Kasparova LSB Dutton Raskutti Sa Morganti
%%  LocalWords:  Emonts Weijmans Ibata McConnachie Schweizer Magain
%%  LocalWords:  Oosterloo baryonic Buote GALEX Salim Kelson outlier
%%  LocalWords:  Bruzual Ahn Willett SQL Chilingarian Saulder Swaters
%%  LocalWords:  Krajn GN Kuntschner situ indices Emsell Capp CasJobs
%%  LocalWords:  Schmitz circ accrete jabref CMD

>>>>

The another formation scenario for GLSB systems is that they are normal galaxies that undergo a high angular momentum minor merger where the LSB disk is the shredded remains of an accreted galaxy.  \citet{Pen06} find that extended disks can be formed via the disruption of dwarf galaxies.  These simulated debris disks have exponential profiles and extend out to 200 kpc from the central galaxy.

Observations of Malin~1 support the accretion model of GLSBs.  \citet{Barth07} analyze HST imaging of Malin~1 and find that the central object is well described as a normal early-type barred galaxy that was previously unresolved in ground-based imaging. \citet{Barth07} points out that extended disk structures in galaxies such as M31 appear to be a smooth continuation of the inner disk \citep{Ferguson02}, while Malin~1's extended disk is better described as a distinct component, (although \citet{Ibata05} claim M31's extended disk is of accretion origin).

If GLSB disks are indeed the remains of accreted dwarf galaxies, they represent a unique opportunity to observationally probe the outer regions of galaxy dark matter halos, especially if they form around elliptical galaxies.  The mass profiles of elliptical galaxies have been difficult to constrain since they lack the extended \HI\ disks found around spirals.  As an example of how difficult it can be to measure the outer regions of elliptical galaxy kinematics, \citet{Murphy2011} model M87 out to $\sim$23 kpc with stellar kinematics and out to 51 kpc with globular clusters (or $R=4.8 R_e$, where $R_e$ is the effective radius). \citet{Raskutti14} are also limited to measuring kinematics out to $\sim$2-5 $R_e$.

@@

\begin{deluxetable*}{l l c c c c c c}
%\tabletypesize{\small \footnotesize \scriptsize}
\tabletypesize{\small}
%\rotate
\tablewidth{0pt}
%\tablenum{num}
%\tablecolumns{num}
%\tableheadfrac{num}
\tablecaption{Basic Properties of GLSB Galaxies}
\tablehead{
\colhead{Name} & \colhead{SDSS ID}   &  \colhead{z}  &\colhead{$r$}  & \colhead{$R_e$} 
& \colhead{$\sigma$}  & \colhead{\HI\ Mass} & \colhead{Stellar Mass} \\
     &  \colhead{(DR 10)}   &     &  \colhead{(mag)}    & \colhead{arcsec} & \colhead{\kms} & \colhead{log$_{10}$M$_{\odot}$} & \colhead{log$_{10}$M$_{\odot}$}
\label{basic_prop}}    
\startdata
Malin 1 & 1237661068728139925 & 0.083 & 16.70 & 1.86 & 177 &  10.9 & 11.1 \\
Malin 2 & 1237667735572381852 &   0.046 & 14.34 & 7.17 & 202  &  10.6 & 11.3 \\
UGC 06614 & 1237668585435365393 &  0.021 & 13.52 & 7.3 & 150 &  10.5 & 10.9
\enddata
\tablecomments{\HI\ masses updated to our adopted cosmological parameters from \citet{Pickering97}.  The rest taken from SDSS. Stellar masses calculated from \citet{Conroy09}}
\end{deluxetable*}

% SDSS stellar masses.  using stellarMassFSPSGranEarlyDust:
% Malin 2:  logMass	11.3307373297902
% Malin 1:  logMass	11.1328074307929
% UGC 6614: logMass	10.8745260407613

@@

\begin{figure*}
\epsscale{.4}
\plotone{Plots/Color_Magnitu_Diagram.eps}\plotone{Plots/Faber_Jackson_Relation.eps}\\
\plotone{Plots/Fundamental_Plane.eps}\plotone{Plots/mg_sigma.eps}
\epsscale{1}
\caption{
Top Left:  The color-magnitude diagram for 141,000 galaxies of similar redshift and magnitude as Malin 2 and UGC 6614.  Top Right:  The Faber-Jackson relation for 45,700 SDSS galaxies, along with a best-fit line.  Left:  The $r$-band fundamental plane from SDSS along with our GLSB galaxies and best-fit line.  Bottom Right:  The relation between central velocity dispersion and the Mg2 absorption feature.   Looking at the SDSS data, our GLSB galaxies appear as relatively typical red elliptical galaxies.  \label{elliptical_relations}}
\end{figure*}

@@

In Figure~\ref{disk_comp}, we compare the sizes of the stellar disks of the GLSB galaxies with other observations.  For this plot, we take the maximum detected radius from the literature for the galaxies.  It is possible that deeper observations could increase the values of $R_{max}$.  While UGC 6614 is close in size to the upper end of local \HI\ disks, Malin 1 and 2 are much larger than galaxies observed in the THINGS survey.  This seems to point against the GLSB formation mechanism of  \citet{Mapelli08}, as we do not observe \HI\ disks large enough to convert into GLSB systems unless an interaction is able to boost the \HI\ angular momentum.

%/astro/store/shared-scratch1/yoachim/Research/GLSB/THINGS/thingsplot.py
\begin{figure*}
\epsscale{.5}
\plotone{Plots/things.eps}
\epsscale{1}
\caption{ Comparison of the GLSB disks with other galaxy disks from the literature.  THINGS data taken from \citet{DeBlok2008}, Malin 1 from \citet{Galaz15}, Malin 2 and UGC 6614 from \citet{Pickering97}.  Here, $R_{max}$ is the maximum radius with an \HI\ detection for the THINGS galaxies, Malin~2, and UGC~6614 and the maximum radius with a stellar flux detection for Malin~1 XXX.  It is interesting to note that the \emph{stellar} disks of GLSB galaxies can be larger than the \HI\ disks of nearby systems.  \label{disk_comp}}
\end{figure*}

@@

We use the kinemetry package presented in \citet{Krajn06} to fit a tilted-ring model to the stellar kinematics.  For these fits, we hold the position angle and inclination fixed to values found from fitting the fiber-photometry.  Figures~\ref{malin_kinem} and~\ref{ugc_kinem} show the kinemetry fits and extracted rotation curves.  We iterate the fit, clipping residuals which deviate from the model by more than 250 \kms.

@@

Following the work in \citet{Yoachim10,Yoachim12}, we generate a suite of model galaxy SEDs using relatively simple star formation histories and find the best fitting single model.  We break the star formation history into three age bins, 0-10 Gyr, 10-11.5 Gyr, and 11.5-12 Gyr.  We assume each age bin has a constant star formation with a level of 0, 0.5, 1, or 10 solar-masses per year.  Each SFH can have a metallicity of [Z/H] = -2.3,-1.70,-0.70, -0.40, 0.0, or  0.40.  We do not include any metallicity evolution, as all the age bins in a given model has the same metallicity.  This gives us 361 unique non-zero model spectra. We use \citet{Bruzual03} with a Chabrier IMF to generate the model spectra.

@@

%--now on new mac.  /Users/yoachim/Research/GLSB/THINGS/guess_projen.py
\begin{figure*}
\epsscale{0.5}
\plotone{Plots/things_projenguess.eps}
\epsscale{1}
\caption{Comparison of GLSB \HI\ masses \citep{Pickering97} with \HI\ masses measured in the THINGS survey \citep{Walter2008}.  While the \HI\ masses in the GLSB systems are high, they are consistent with what would be available in a merger with a V$_{\rm{max}}\sim$110 km/s or smaller galaxy, assuming the merger did not trigger large amounts of star formation and feedback. \label{thingsMHI}}
\end{figure*}

@@

While the old stellar populations in the disks is evidence against the \citet{Mapelli08} formation scenario, we have not addressed what collisional ring galaxies do evolve into.  Since modern photometric surveys are often surface brightness limited, it is possible a gLSB system will be found in the future that is descended from a collisional ring.

@@

We have observed two giant low surface brightness galaxies with the VIRUS-P IFU.  We find the systems are best explained as massive elliptical galaxies which have managed to accrete stars and gas into a large disk.  We find the inner bulge is similar to elliptical galaxies found in SDSS.  The stars have a disk-like velocity field, are dominated by old stars, but do show signs of ongoing star formation. The stars also show a sharp metallicity transition, consistent with the disk forming independently of the bulge. The \HI content of the disk is also consistent with coming from a low-mass accreted galaxy.

@@

\appendix
\section{SDSS SQL Queries} \label{SQL}
To build our sample of comparison galaxies from SDSS, we used the CasJobs interface to query the SDSS DR10 database.

@@

% LocalWords:  Debattista Horrocks Lancashire RCUK Malin Stanonik Spavone Lelli
% LocalWords:  Bothun GLSB Das Mapelli Weij Loebman SungWon Kwak Scd Sd
% LocalWords:  Schiminovich Iodice Belokurov Fairchild RMS pPXF Cappellari th
% LocalWords:  GANDALF Sarzi CosmoloPy Bernardi GLSBs DeBlok Verheijen Yoachim
% LocalWords:  SEDs Brookhaven Instituto Astrofisica Canarias Notre JINA DM
% LocalWords:  IFU UGC HSB SAURON ellipticals Spitzer IC Galaz
% LocalWords:  kpc HIPASS CCD metallicity MANGA SDSS arcsec lensing IRAF
% LocalWords:  Spectrograph spectrograph IDL longslit photometric FWHM SSP mis
% LocalWords:  dispersions Balmer decrement fiducia redshift Chabrier
% LocalWords:  Faber redshifts NGC kinemetry photometry anisotropy Gyr SFH
% LocalWords:  metallicities colliders rotators et al MNRAS de datacube Ppxf
% LocalWords:  dashed CDM Kasparova LSB Dutton Raskutti Sa Morganti
%%  LocalWords:  Emonts Weijmans Ibata McConnachie Schweizer Magain
%%  LocalWords:  Oosterloo baryonic Buote GALEX Salim Kelson outlier
%%  LocalWords:  Bruzual Ahn Willett SQL Chilingarian Saulder Swaters
%%  LocalWords:  Krajn GN Kuntschner situ indices Emsell Capp CasJobs
%%  LocalWords:  Schmitz circ accrete jabref CMD
